
\section{Future work}

\noindent \textsc{Shorter term projects}

\subsection{Multisimplicial cup product}

Following a suggestion by P.~May, with A.~Pizzi and P.~Salvatore, we define in upcoming work an $E_\infty$-structure on multisimplicial cochains \cite{medina2021multisimplicial}.
This structure generalizes the $E_\infty$-structures defined on simplicial and cubical cochains discussed in \cref{ss:e-infty structures}.

\subsection{Cochain level secondary operations}

Using the formulas of \cite{medina2021maysteenrod} for cochain level Steenrod operations over all primes.
I will extend our work constructing cochains that enforce the Cartan and Adem relations of Steenrod operations with mod $2$ coefficients (\cref{ss:cartan} - \cref{ss:adem}) to all primes.
The case of the Cartan relation is currently being written, whereas that of the Adem relation requires new ideas.

\subsection{Free loop spaces and string topology}

In \cite{medina2021cobar} we provided an $E_\infty$-model for the based loop space of non-necessarily simply-connected spaces.
In future work we will obtain a similar model for the chains on the free loop space, which may be thought of as a twisted tensor product of two $E_\infty$-coalgebras.
Then we will incorporate Poincar\'{e} duality and try to recover string topology in the non-simply connected case with all of its subtleties.

\subsection{Representation and cohomology of categories}

The theory of representations and cohomology of groups can be recasted in terms of (Hopf) group rings.
Many of the results in this theory extend naturally to quivers and (small) categories but not all.
In particular, some important functoriality properties are lost in all current generalizations from groups to categories.
In work to appear, I argue that to prevent this loss one should consider the $E_2$-coalgebra structure carried by the chains on the nerve of the category.
When the category is a group, this corresponds to considering the $E_2$-coalgebra structure on the bar construction of the group algebra.

\vspace*{6pt} \noindent \textsc{Longer Term Projects}

\subsection{Geometric cohomology} \label{ss:geometric cohomology}

As part of a long term project with D. Sinha and G. Friedman we are working to provide a firm foundation for a model of ordinary (co)homology of smooth manifolds based on maps from manifolds with corners as generating (co)chains \cite{medina2021foundations}.
The first application of this theory already appeared in \cite{medina2021flowing}, where we use a vector field flow defined through a cubulation of a closed manifold to reconcile the partially defined commutative product on geometric cochains with the standard cup product on cubical cochains, which is fully defined and commutative only up to coherent homotopies.
Several more applications are to follow.

\subsection{Explicit Lie models for connected spaces} \label{ss:lie models}

Over $\Q$, a chain approximation to the diagonal can be symmetrized, giving rise to a cocommutative coalgebra structure on cellular chains.
This coalgebra cannot be made simultaneously coassociative, but this relation can be imposed in a derive sense through a family of coherent chain homotopies -- which also respect certain symmetry constrains -- and give rise to a so called $C_\infty$-coalgebra structure.
One can think of $C_\infty$-coalgebras in terms of the somewhat more familiar notion of $A_\infty$-coalgebra where cocommutativity is satisfied strictly.
As a manifestation of Koszul duality, a $C_\infty$-coalgebra structure on cellular chains is equivalent to a differential on the completion of the free graded Lie algebra generated by the cells shifted downwards in degree by one.
This relates $C_\infty$-coalgebras to deformation theory through Deligne's principle.
For cell complexes whose closed cells have the $\Q$-homology of a point, Dennis Sullivan provided in \cite{sullivan2007appendix} a local inductive construction defining a $C_\infty$-coalgebra structure on their cellular chains.
We reprint a challenge he posted regarding the resulting structure.
\begin{displaycquote}[p.2]{lawrence2014interval}
	\textsc{Problem}. Study this free differential Lie algebra attached to a cell complex, determine its topological and geometric meaning as an intrinsic object.
	Give closed form formulae for the differential and for the induced maps associated to subdivisions.
\end{displaycquote}
As proven by Quillen, the quasi-isomorphism type of this $C_\infty$-coalgebra is a complete invariant of the rational homotopy type of simply-connected spaces.
For the $C_\infty$-coalgebra structure on the interval, Dennis and Ruth Lawrence addressed the challenge reprinted above introducing a formula for it which can be interpreted in terms of parallel transport of flat connections \cite{lawrence2014interval}, and for which the subdivision map is described by the Baker--Campbell--Hausdorff formula.

To generalize Quillen's equivalence of homotopy categories to one between (not necessarily 1-connected) simplicial sets Buijs, F{\'e}lix, Murillo, and Tanr{\'e} provided in \cite{buijs2020liemodels} an axiomatic characterization of a $C_\infty$-coalgebra structures on the chains of standard simplices extending the Lawrence--Sullivan structure.
Their construction, also presented independently by Bandiera and Robert-Nicoud using operads, agrees after linear dualization with the one obtained by Cheng and Getzler in \cite{getzler2008transfering}, where they showed that the Kontsevich--Soibelman sum-over-trees formula defining the transfer of $A_\infty$-algebras through a chain contraction induces a transfer of $C_\infty$-algebras.
This allowed them to construct a $C_\infty$-algebra structure on simplicial cochains by transferring Sulivan' polynomial differential forms \cite{sullivan1977infinitesimal} through Dupont's contraction.
The resulting description is given in terms of rooted trees.

$C_\infty$-coalgebras are controlled by the operad $\com_\infty$ which is the Koszul resolution of the operad $\com$, i.e., the cobar construction applied to the suspension of the cooperad $\lie^{\mathrm c}$, the Koszul dual cooperad of $\com$.
Another interesting resolution of $\com$ is constructed concatenating the bar and cobar constructions.
This resolution method is an algebraic version of the $W$-construction of Boardman--Vogt.
In \cite{vallette2020higherlietheory}, Daniel Robert-Nicoud and Bruno Vallette studied coalgebras over this resolution of $\com$, which they termed $CC_\infty$-coalgebras.
They constructed on the chain of standard simplices natural $CC_\infty$-coalgebra structures and described them explicitly using bicolored trees.

Despite the explicit constructions reviewed above, the problem quoted earlier related to closed form formulae remains open.
One possible avenue to generalize to cubical chains the formula defining the Lawrence--Sullivan $C_\infty$-coalgebra on $\gchains(\gcube)$, is to define the tensor product of $C_\infty$-coalgebras and then extend it monoidally to all cubes via the isomorphism $\chains(\cube^n) \cong \gchains(\gcube)^{\ot n}$.
The monoidal structure on the category of $A_\infty$-coalgebras is defined through a chain approximation to the diagonal of the Stasheff polytopes compatible with the operad structure, a diagonal interpreted geometrically by Masuda--Thomas--Tonks--Vallette in \cite{vallette2021associahedra}.
Unfortunately, the resulting $A_\infty$-coalgebra on $\gchains(\gcube)^{\ot 2}$ is not $C_\infty$.
This could be corrected through an algebraic symmetrization of the associahedral diagonal, and deeper understanding of the space of diagonals of polytopes.

\subsection{Fiber polytopes and Steenrod diagonals} \label{ss:polytopes}

Together with Guillaume Laplante-Anfossi we are exploring the geometry of higher diagonals of convex polytopes.
This work connects to higher category in the form pioneered by Kapranov and Voevodsky \cite{kapranov1991polycategories}.
We have already achieved substantial progress in the study of higher Steenrod diagonals using the notion of fiber polytopes \cite{billera1992fiber.polytopes}.
We introduced a generalization of the Masuda--Thomas--Tonks--Vallette construction of a cellular diagonal from a admissible vector, to the construction of a Steenrod diagonal from an admissible orthonormal frame.
This is a very exciting connection bringing convex geometry, higher categories, and homotopy theory closer together, from which a vast area of research opens.
Several projects are growing from this connection whose completion will involve Ralph Kaufmann, Guillaume Laplante-Anfossi, Viktoriya Ozornova, Bruno Vallete and other researchers.

\subsection{Odd prime Steenrod operations: persistence theory and Khovanov homology} \label{ss:odd prime steenrod operations}

In work being finalized, with Federico Cantero-Mor\'{a}n we introduced a description of Steenrod operations at the simplicial cochain level that is dual to the one given in \cite{medina2021maysteenrod}.
This is a generalization of the description in \cite{medina2021newformulas} of cup-$i$ products to cup-$(p,i)$ products.
There are two sources of motivation for this.
On one hand, for applications in persistent homology or computational topology more broadly, this description lends itself to a much faster algorithmic implementation.
On the other, this dual description allowed Cantero-Mor\'an to define Steenrod operations at the even prime on Khovanov homology \cite{cantero-moran2020khovanov}.
We will use our dual description of cup $(p,i)$-products to define said operations on Khovanov homology at odd primes as well.
This project will involved several graduate students and other young researcher, for which some financial support has already been secured.

\subsection{Homotopical algebraic data analysis and AI} \label{ss:ai}

Topology can complement traditional approaches to machine learning and data analysis by providing global summaries of complex relational structures.
The added information is model-independent and highly resistant to noise.
Designing reliable AI systems is a multi-faceted challenge and homotopical algebra can be used to tackle the following core aspects of the problem:

\begin{itemize}
	\item Robustness to noise and adversarial attacks: deep learning models are surprisingly susceptible to small perturbations, and this can lead to unexpected failures in deployed systems.
	\item Generalizability to unseen data: large models can overfit the data they are trained on and thus fail to generalize to data in deployment.
\end{itemize}
Traditional regularizers, one of the primary tools used to tackle the above challenges, impose penalties on representations that exhibit undesirable properties such as exploding gradients or highly complicated geometric structures.
By studying the topology of the weight space and decision boundaries of trained models, we will validate on real-world data the assumption that robustness is linked to less cumbersome topologies.
This will enable to create performant topological regularizers that improve the resilience and generalization power of new models.

To quantify the topological complexity of these spaces and device such regularization procedures, developments from the sub-field of homotopical algebraic data analysis, such as Steenrod persistence modules, are fundamental.
Together with part of \texttt{giotto-tda}'s team, we will develop -- at all stages from theory and algorithms to implementation and deployment -- novel homotopical invariants that can be used for these and similar challenges.
International collaborations are also envision, in particular, with the Laboratory for Topology and Neuroscience at EPFL, where as a member I wrote the mathematical content of a grant on this topic awarded for over 1 million USD.

\subsection{Symmetry protected topological phases and cochain constructions} \label{ss:spt phases}

A central problem in physics is to define and understand the moduli ``space'' of quantum systems with a fixed set of invariants, for example their dimension and symmetry type.
In condensed matter physics, quantum systems are presented using \textit{lattice models} which, intuitively, are given by a Hamiltonian presented as a sum of local terms on a Hilbert space associated to a lattice in $\R^n$.
We think of these as defined on flat space.
One such system is said to be \textit{gapped} if the spectrum of the Hamiltonian is bounded away from $0$, and two Hamiltonians represent the same \textit{phase} if there exists a deformation between them consisting only of systems that remain bounded from below.

Given a lattice model, using cellular decompositions and state sum type constructions one can often compute the corresponding \textit{partition functions} on spacetime manifolds from actions expressed in terms of gauge fields represented by cochains and cochain level structures: Stiefel--Whitney cochains, cup-$(p, i)$ products (\cref{ss:cup-i}, \cref{ss:may steenrod}) and Cartan/Adem coboundaries (\cref{ss:cartan}, \cref{ss:adem}) among others.
Subdivision invariance gives rise to a functorial TQFT, which in the \textit{invertible} case is expected to be controlled by a generalized cohomology theory \cite{freed2021reflection}.
The cochain level structure used in the definition of the cellular gauge theory is interpreted from this point of view as describing a cochain model of the Postnikov tower of the relevant spectrum.
For example, fermionic phases protected by a $G$-symmetry are believed to be classified by applying to $BG$ the Pontryagin dual of spin bordism \cite{kapustin2015cobordism, kapustin2017fermionic}.
Building on these insights and using a formula of mine \cite{medina2020cartan}, A. Kapustin proposed a structural ansatz in low dimensions that Greg Brumfiel and John Morgan verified by constructing cochain models of certain connective covers of said spectrum \cite{brumfiel2016pontrjagin, brumfiel2018pontrjagin}.

I will deepen the understanding of SPT phases by further connecting these point of view: lattice models, discrete gauge theories, stable homotopy theory, functorial field theories.
This is a long term project, but a first concrete objective is the description of the signature of 4 manifolds through a combinatorial local state sum formula.
The importance of this question has been highlighted by Peter Teichner.
Partial results for general $4k$-manifolds have been achieved by Dennis Sullivan and Andrew Ranicki \cite{sullivan1976signature}.


%%%%%%%%%%%%%%%%%%%5

%\subsection{Foundations of geometric (co)homology} \label{ss:foundations}
%
%The use of manifold-like objects to represent homology and cohomology classes has a long history and many technical subtleties.
%In \cite{medina2021foundations}, with G. Friedman and D. Sinha we further develop one such theory for smooth manifolds based on maps from manifolds with corners.
%As an application of these foundations, we expect to extend our work relating transverse intersection and the cup product of cubical cochains (\cref{ss:flows}) to an entire $E_\infty$-structure.
%More explicitly, we expect to introduce further coherent vector fields to model geometrically the $E_\infty$-structure defined in \cite{medina2021cubical}.

%\subsection{Topological lattice field theory}
%
%I organize a seminar at the Max Planck Institute exploring the connections between combinatorial topology and topological quantum field theory.
%Given that Steenrod \mbox{cup-$i$} products and their Adem and Cartan coboundaries feature centrally on certain field theories \cite{gaiotto2016spin, kapustin2017fermionic, barkeshli2021classification}, we expect to use the newly introduced cup-$(p,i)$ products and their Cartan and Adem cochains in the construction of new theories.
%
%\subsection{Higher categories}
%
%In higher category theory, the paper \cite{medina2020globular} constructs a functor producing strict $\infty$-categories from group-like cup-i coalgebras.
%In particular, the cup-$i$ constructions on simplices and cubes define, respectively, the Street nerve and cubical nerve of strict $\infty$-categories.
%We anticipate that the more general Steenrod $(p,i)$-products constructed in \cite{medina2021maysteenrod} will also have deep combinatorial interpretations in category theory which we are currently exploring.
%
%\subsection{Dyer-Lashof operations}
%
%Configuration space of $r$ points in $\mathbb R^n$ are objects of central topological interest related to loop spaces, knots, and deformation theory.
%Recently, they have also found applications in motion planning and robotics through the concept of topological complexity \cite{farber2003motion.planning}.
%The homology of these spaces is a crucial invariant and we expect to identify chain level representatives for these classes on combinatorial models of the $E_n$-operads.
%This work is based on constructions in the $E_\infty$ case discussed in \cref{ss:may steenrod}.
%
%
%\subsection{\pdfEinfty-algebras and $L$-theory}
%
%With Manuel Rivera and Mahmoud Zeinalian we expect to combine the algebraic structures describing homotopy types \cite{sullivan1977infinitesimal, mandell2001padic} and those developed by Ranicki \cite{ranicki1992topological} for the classification of topological manifold structures.
%Substantial partial results in this direction can be found in their paper \cite{rivera2018rigidification} and my thesis \cite{medina2015thesis}.
%
%\subsection{Explicit Lie models for connected spaces}
%
%After the pioneering work of Quillen \cite{quillen1969rational} and Sullivan \cite{sullivan1977infinitesimal}, it is known that the categories of differential Lie and commutative (co)algebras model the rational homotopy category of spaces under certain finiteness assumptions.
%In \cite{lawrence2014interval} R. Lawrence and D. Sullivan define an explicit model of the interval which was extended non-constructively in \cite{buijs2020liemodels} to all simplices.
%I expect to provide a explicit formula of models for all cubes using a symmetrization of the associahedral diagonal.
%
%\subsection{Symmetries of the associahedral diagonal}
%
%Defining cellular diagonals of polytopes is a problem with combinatorial interest.
%In the case of Stasheff's associahedra, polytopes controling $A_\infty$-algebras,
%the existence of a diagonal ensures that the category of $A_\infty$-algebras is equipped with a well defined tensor product.
%This structure is key for applications in Heegard-Floer homology and other areas of symplectic geometry via Fukaya categories \cite{lipshitz2020diagonals}.
%
%Using my model of the $E_\infty$-operad, I expect to introduce a family of coherent homotopies correcting the lack of symmetry of the associahedral diagonal and other polytopes of interest.
