
\section{Future work}

\subsection{Topology and deep learning}

Building on the work by \texttt{giotto-tda}'s team the following research direction will be pursued.
It already counts with a public-private Innosuisse grant approved for over 1 Million USD, whose mathematical content I wrote as part of Kathryn Hess' lab.
Topology can complement traditional approaches to machine learning and data analysis by providing global summaries of complex relational structures.
The added information is model-independent and highly resistant to noise.
Designing reliable AI systems is a multi-faceted challenge and topology can be used to tackle the following core aspects of the problem:
\begin{itemize}
	\item Robustness to noise and adversarial attacks: deep learning models are surprisingly susceptible to small perturbations, and this can lead to unexpected failures in deployed systems.
	\item Generalizability to unseen data: large models can overfit the data they are trained on and thus fail to generalize to data in deployment.
\end{itemize}
Traditional regularizers, one of the primary tools used to tackle the above challenges, impose penalties on representations that exhibit undesirable properties such as exploding gradients or highly complicated geometric structures.
By studying the topology of the weight space and decision boundaries of trained models, we will validate on real world data the assumption that robustness is linked to less cumbersome topologies.
This will enable to create performant topological regularizers that improve the resilience and generalization power of new models.

\subsection{Foundations of geometric (co)homology} \label{ss:foundations}

The use of manifold-like objects to represent homology and cohomology classes has a long history and many technical subtleties.
In \cite{medina2021foundations}, with G. Friedman and D. Sinha we further develop one such theory for smooth manifolds based on maps from manifolds with corners.
As an application of these foundations, we expect to extend our work relating transverse intersection and the cup product of cubical cochains (\cref{ss:flows}) to an entire $E_\infty$-structure.
More explicitly, we expect to introduce further coherent vector fields to model geometrically the $E_\infty$-structure defined in \cite{medina2021cubical}.

\subsection{Steenrod operations and Khovanov homology}

In work being finalized, with Federico Cantero-Mor\'{a}n we introduced a description of Steenrod operations at the simplicial cochain level that is dual to the one given in \cite{medina2020maysteenrod}.
This is a generalization of the description in \cite{medina2021newformulas} of cup-$i$ products to cup-$(p,i)$ products.
There are two sources of motivation for this.
On one hand, for applications in persistent homology or computational topology more broadly, this description lends itself to a much faster algorithmic implementation.
On the other, this dual description is applicable in more generality.
For example, the dual description of \cite{medina2021newformulas} was used by Cantero-Mor\'{a}n to define Steenrod squares on Khovanov homology \cite{cantero-moran2020khovanov}, and we expect to define Steenrod operations at odd primes using our new description.

\subsection{Cochain level secondary operations}

Building on our work described in \cref{ss:cartan} and \cref{ss:adem} constructing cochains that enforce the Cartan and Adem relations of Steenrod operations with mod $2$ coefficients, we will generalize these, using the formulas of \cite{medina2020maysteenrod}, to Steenrod operations over all primes.
This will allow for the effective description of secondary operations and the approximation via chain complexes of certain Postnikov towers, as done for the Spin bordism spectrum in \cite{brumfiel2016pontrjagin, brumfiel2018pontrjagin}.

\subsection{Topological lattice field theory}

I organize a seminar at the Max Planck Institute exploring the connections between combinatorial topology and topological quantum field theory.
Given that Steenrod \mbox{cup-$i$} products and their Adem and Cartan coboundaries feature centrally on certain field theories \cite{gaiotto2016spin, kapustin2017fermionic, barkeshli2021classification}, we expect to use the newly introduced cup-$(p,i)$ products and their Cartan and Adem cochains in the construction of new theories.

\subsection{Multisimplicial cup product}

Following a suggestion by P.~May, with A.~Pizzi and P.~Salvatore, we define in upcoming work an $E_\infty$-structure on multisimplicial cochains \cite{medina2021multisimplicial}.
This structure generalizes the $E_\infty$-structures defined on simplicial and cubical cochains discussed in \cref{ss:e-infty structures}.

\subsection{Higher categories}

In higher category theory, the paper \cite{medina2020globular} constructs a functor producing strict $\infty$-categories from group-like cup-i coalgebras.
In particular, the cup-$i$ constructions on simplices and cubes define, respectively, the Street nerve and cubical nerve of strict $\infty$-categories.
We anticipate that the more general Steenrod $(p,i)$-products constructed in \cite{medina2020maysteenrod} will also have deep combinatorial interpretations in category theory which we are currently exploring.

\subsection{Dyer-Lashof operations}

Configuration space of $r$ points in $\mathbb R^n$ are objects of central topological interest related to loop spaces, knots, and deformation theory.
Recently, they have also found applications in motion planning and robotics through the concept of topological complexity \cite{farber2003motion.planning}.
The homology of these spaces is a crucial invariant and we expect to identify chain level representatives for these classes on combinatorial models of the $E_n$-operads.
This work is based on constructions in the $E_\infty$ case discussed in \cref{ss:may steenrod}.

\subsection{Representation and cohomology of categories}

The theory of representations and cohomology of groups can be recasted in terms of (Hopf) group rings.
Many of the results in this theory extend naturally to quivers and (small) categories but not all.
In particular, some important functoriality properties are lost in all current generalizations from groups to categories.
In work to appear, I argue that to prevent this loss one should consider the $E_2$-coalgebra structure carried by the chains on the nerve of the category.
When the category is a group, this corresponds to considering the $E_2$-coalgebra structure on the bar construction of the group algebra.

\subsection{Free loop spaces and string topology}

In \cite{medina2021cobar} we provided an $E_\infty$-model for the based loop space of non-necessarily simply-connected spaces.
In future work we will obtain a similar model for the chains on the free loop space, which may be thought of as a twisted tensor product of two $E_\infty$-coalgebras.
Then we will incorporate Poincar\'{e} duality and try to recover string topology in the non-simply connected case with all of its subtleties.

\subsection{\pdfEinfty-algebras and $L$-theory}

With Manuel Rivera and Mahmoud Zeinalian we expect to combine the algebraic structures describing homotopy types \cite{sullivan1977infinitesimal, mandell2001padic} and those developed by Ranicki \cite{ranicki1992topological} for the classification of topological manifold structures.
Substantial partial results in this direction can be found in their paper \cite{rivera2018rigidification} and my thesis \cite{medina2015thesis}.

\subsection{Explicit Lie models for connected spaces}

After the pioneering work of Quillen \cite{quillen1969rational} and Sullivan \cite{sullivan1977infinitesimal}, it is known that the categories of differential Lie and commutative (co)algebras model the rational homotopy category of spaces under certain finiteness assumptions.
In \cite{lawrence2014interval} R. Lawrence and D. Sullivan define an explicit model of the interval which was extended non-constructively in \cite{buijs2020liemodels} to all simplices.
I expect to provide a explicit formula of models for all cubes using a symmetrization of the associahedral diagonal.

\subsection{Symmetries of the associahedral diagonal}

Defining cellular diagonals of polytopes is a problem with combinatorial interest.
In the case of Stasheff's associahedra, polytopes controling $A_\infty$-algebras,
the existence of a diagonal ensures that the category of $A_\infty$-algebras is equipped with a well defined tensor product.
This structure is key for applications in Heegard-Floer homology and other areas of symplectic geometry via Fukaya categories \cite{lipshitz2020diagonals}.

Using my model of the $E_\infty$-operad, I expect to introduce a family of coherent homotopies correcting the lack of symmetry of the associahedral diagonal and other polytopes of interest.
